Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-06-05 paper.tex
Fixes errors and adds conclusions

But this of course presupposes that the affective state of the student
is available to the system. This detection and recognition can be done
in real time in different ways. In the wider context of the detection
of emotions and affective 
state, many researchers have
demonstrated that affect-aware computers can provide a better performance in
assisting humans
\cite{picard2001toward,munson2018models,bosch2017affective}, and that
providing a richer experience while learning to program might improve
the outcome by making the students more engaged with it
\cite{asensio2014progamer}. However, when ILEs embrace these methodologies
and techniques a
problem arises: Sensors which are not present in the default computer
configuration must be used to gather data that will be used to perform the
recognition of users\' affective states.
Some sensors rely on physiological readings
and must be in physical contact with students \cite{atkinson2017assessing,swansianalyzing}.
These sensors can be considered
intrusive or invasive, not to mention that you have to acquire them
and convince the student to wear them while learning.  They can also disrupt the student\'s learning experience
\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
Other less cumbersome sensors, such as video cameras, which are in
fact default equipment in most laptops, can also be considered invasive, since
they always transmit the user's identity, appearance and behavior along with
the emotional data \cite{picard2001toward}. So what we would need in
order to embed recognition of affective states in learning systems
would be non-invasive, anonymous and also ubiquitous sensors. In this
paper we are going to test such a kind of system.

@@

The aim of this research was to evaluate the use of keyboard and mouse
dynamics (KD) as an appropriate sensory input for an affective
recognition system for students of different levels learning to
program, specifically carrying out assignments consisting in writing
short programs interactively. An experimental approach was adopted
with this objective, by effectively using the method with real
subjects using the environment and using KD measurements and survey
responses as training data for several machine learning systems.

The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ measure below their counterparts of
fixed-text; however, in this
work the results obtained were similar to those works that worked with fixed-text
dynamics. A possible explanation of this would be the addition of the mouse
dynamics features and the additional preprocessing performed on the
feature vectors. Out of all methods tested, tree induction algorithms
obtained better results when we take into consideration both accuracy
and $\kappa$ values. These methods, besides, are able to give us some
insight on what are the KD features that effectively describe the
different affective states.

While these are promising results, further experiments are
needed. Other experiments should focus on novice programmers, with
which we expect to gather more data and more representative examples for all the
classes. These experiments should give researchers the ability to do stratified samples to balance
the class distribution and achieve better results. Also more experiments should
give more insight on the relationship between affective state  and interactive
behavior. Results show near 80\% accuracy, these results  are competitive
against other classifiers of a similar domain, but we do not know yet what is
the effectiveness when applied to affective-aware learning environments.

%This
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text?
% Is 80% effective?
%  - done  Mario 
Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
\cite{blikstein2011using}; the parameters proposed by this author
include the size of the program and changes in the code. Other authors
also propose treating the use of certain keys as features e.g., the
delete key because it is mainly used for correcting sentences. Also a
multi-modal approach is needed, using attention detectors alongside KD
dynamics to measure at least the boredom/attention state independently
of the verbal assessment by the student. Since all these need a
completely new approach, a new methodology will be developed to 
gather data and take measures.

>>>>

But this of course presupposes that the affective state of the student
is available to the system. This detection and recognition can be done
in real time in different ways. In the wider context of the detection
of emotions and affective 
state, many researchers have
demonstrated that affect-aware computers can provide a better performance in
assisting humans
\cite{picard2001toward,munson2018models,bosch2017affective}, and that
providing a richer experience while learning to program might improve
the outcome by making the students more engaged with it
\cite{asensio2014progamer}. However, when ILEs embrace these methodologies
and techniques a
problem arises: Sensors which are not present in the default computer
configuration must be used to gather data that will be used to perform the
recognition of users\' affective states.
Some sensors rely on physiological readings
and must be in physical contact with students \cite{kula2017assessing,swansianalyzing,gonzalez2017affect}.
These sensors can be considered
intrusive or invasive, not to mention that you have to acquire them
and convince the student to wear them while learning.  They can also disrupt the student\'s learning experience
\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
Other less cumbersome sensors, such as video cameras, which are in
fact default equipment in most laptops, can also be considered invasive, since
they always transmit the user's identity, appearance and behavior along with
the emotional data \cite{picard2001toward}. So what we would need in
order to embed recognition of affective states in learning systems
would be non-invasive, anonymous and also ubiquitous sensors. In this
paper we are going to test such a kind of system.

@@

The aim of this research was to present an affective recognition
system that uses keyboard and mouse
dynamics (KD) as a non-intrusive sensory input. This system was
applied to  for students of different levels learning to
program by carrying out assignments consisting in writing
short programs interactively. An experimental approach was adopted
with this objective, by effectively taking KD measurements on real
subjects using the environment; these subjects they responded to
surveys. Measurements and surveys were merged into a dataset that was
used for training different machine learning system, with our
objective being to find out which method yields the best results in
this environment.

Results obtained with these machine learning systems show that the
method used for gathering affective data convey sufficient information
for an adequate classification of affective states; in fact, although
it is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ measure below their counterparts of
fixed-text; in this work the classification accuracy obtained  with
the best machine learning systems were similar to those using
fixed-text. This level of accuracy obtained beyond the state of the
art in free-text systems might be due to the addition of
the mouse dynamics features and the additional preprocessing performed on the
feature vectors.

With data captured and preprocessed as indicated in the methodology,
and out of all the machine learning algorithms tested, tree induction algorithms
obtained better results when we take into consideration both accuracy
and $\kappa$ values. These methods, besides, are able to give us some
insight on the key KD features that effectively describe the
different affective states.

While these are promising results, further experiments are
needed. In this work we have used programmers with different levels of
expertise; however, it would be interesting to focus on novice programmers, with
which we expect to gather more data and more representative examples for all the
classes. These experiments should give researchers the ability to do stratified samples to balance
the class distribution and achieve better results. Additional experiments should
give more insight on the relationship between affective state and interactive
behavior. Results show near 80\% accuracy, these results  are competitive
against other classifiers of a similar domain, but we do not know yet what is
the effectiveness when applied to affective-aware learning environments.

A different line of work will focus on extracting other features from
the learners\' interaction such as behavioral data, as proposed by
Blikstein 
\cite{blikstein2011using}; the parameters proposed by this author
include the size of the program and changes in the code. Other authors
also propose treating the use of certain keys as features e.g., the
delete key because it is mainly used for correcting sentences. Also a
multi-modal approach to the detection of the affective state might be
needed, using attention detectors alongside KD 
dynamics to measure at least the boredom/attention state independently
of the verbal assessment by the student. Since all these need a
completely new approach, a new methodology will be developed to 
gather data and take measures.
